---
title: "Zanudo_META_profiles"
author: "Jonas BEAL"
date: "11/17/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Paths
original_path <- normalizePath(getwd())
base_folder <- dirname(dirname(original_path))

knitr::opts_knit$set(root.dir = base_folder)
futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")
```

## Objectives

Generate patients profiles from BRCA METABRIC data for further logical modelling.

```{r packages_and_utilities, echo=FALSE, message=FALSE}

if (!require("pacman")) install.packages("pacman")
list.of.packages <- c("plyr","tidyverse","VennDiagram","networkD3","knitr","gridExtra","mclust","diptest","moments","magrittr")
pacman::p_load(list.of.packages, character.only = TRUE)

tbl_to_df <- function(x){
  x <- x %>% as.data.frame %>%  remove_rownames %>% column_to_rownames(var="PATIENT_ID")
  } 
rename <- dplyr::rename
select <- dplyr::select
```

# Independant omics profiles

## META dataset recap

More than 1800 patients with several kinds of omics data: exome-sequencing, Copy Number Alterations (CNA), RNA and clinical annotations

```{r META_import, echo=FALSE, message=FALSE, warning=FALSE}

#First import META clinical data for further use to correlate with logical modelling results
META_clin <- read_delim("Data/METABRIC/data_clinical_supp_patient.txt", delim = "\t") %>% select(PATIENT_ID, OS_STATUS, OS_MONTHS, CLAUDIN_SUBTYPE) %>% rename(PAM50=CLAUDIN_SUBTYPE) %>% mutate(PAM50=factor(PAM50))

#Due to inconsistencies in genes names (DEC1 replaced by 1-Dec for instance), Entrez ID are used to recover proper HUGO names. Import HUGO/Entrez table for later qualiity check
#Some tables have to be transposed
HUGO_Entrez <- read_delim("Data/Common/HUGO_Entrez.txt", delim = "\t") %>% na.omit
entrez_to_hugo <- function(df_input){
    df_input <- df_input %>% mutate(Hugo_Symbol=unlist(map2(.$Hugo_Symbol,.$Entrez_Gene_Id, function(x,y) if(y %in% HUGO_Entrez$`Entrez Gene ID` & !(y %in% .$Entrez_Gene_Id[duplicated(.$Entrez_Gene_Id) | duplicated(.$Entrez_Gene_Id, fromLast = T)])) HUGO_Entrez$`Approved Symbol`[which(HUGO_Entrez$`Entrez Gene ID`==y)] else x ))) 
    dupli <- duplicated(df_input$Hugo_Symbol) | duplicated(df_input$Hugo_Symbol, fromLast = T)
    df_input$Hugo_Symbol[dupli] <- paste(df_input$Hugo_Symbol[dupli],df_input$Entrez_Gene_Id[dupli],sep = "-")
    df_output <- df_input %>% select(-Entrez_Gene_Id) %>% select(-matches("^X[0-9]+$"))
}

tibble_transpose <- function(df_input){
  df_output <- df_input %>% gather(var, value, -Hugo_Symbol) %>%
    spread(Hugo_Symbol, value) %>%
    rename(PATIENT_ID=var) %>% 
    type_convert
}

#Import META mutations data
META_mut <- read_delim("Data/METABRIC/data_mutations_extended.txt", delim = "\t", skip = 2) %>% select(Tumor_Sample_Barcode, Hugo_Symbol,Variant_Classification, `MA:protein.change`, SIFT, PolyPhen)  %>% rename(PATIENT_ID=Tumor_Sample_Barcode)

META_mut_patients <- readLines("Data/METABRIC/data_mutations_extended.txt",n=1) %>% strsplit(split=" ") %>% unlist %>% tail(length(.)-1)

#Import and process other META omics data
META_CNA <- read_delim("Data/METABRIC/data_CNA.txt", delim = "\t") %>% entrez_to_hugo %>% tibble_transpose
META_RNA <- read_delim("Data/METABRIC/data_expression.txt", delim = "\t") %>% entrez_to_hugo %>% tibble_transpose

#Additional imports: PAM50 gene list
PAM50 <- read_delim("Data/Common/pam50_centroids.txt", delim = "\t") %>% rename(Gene=X1) %>% select(Gene)

#Genes involved in Zanudo model
genenames <- read.table("Models/Zanudo/Zanudo_namesToHugo_curated.txt",header=T,sep="\t")
geneindex <- strsplit(as.character(genenames[,2]), split = ",") %>% sapply(function(l){gsub(" ","",l)})
geneindex <- data.frame(V1 = rep(genenames[,1], sapply(geneindex, length)), V2 = unlist(geneindex))
model_nodes_HUGO <- unique(geneindex[,2]) %>% sub("^\\s+", "", .)

#Create new variables with only genes related to the model
METAmodel_mut <- META_mut %>% filter(Hugo_Symbol %in% model_nodes_HUGO)
METAmodel_CNA <- META_CNA %>% select(PATIENT_ID, one_of(model_nodes_HUGO))
METAmodel_RNA <- META_RNA %>% select(PATIENT_ID, one_of(model_nodes_HUGO))
```

Most patients have all omics data:

```{r VennMETA, echo=FALSE, eval=TRUE}
flog.threshold(ERROR)

#Venn diagramm
grid.draw(venn.diagram(list(CNA=META_CNA$PATIENT_ID, RNA=META_RNA$PATIENT_ID, Mut=META_mut_patients), filename=NULL, col = "transparent", fill = c("red", "blue", "green"), alpha = 0.5, cex = 0.7, fontface = "bold", cat.default.pos = "text", cat.col = c("darkred", "darkblue", "darkgreen"), cat.cex = 1.5, cat.dist = c(0.06, 0.06, 0.03), cat.pos = 0, main = "Patients and available data types"))

grid.newpage()
grid.draw(venn.diagram(list(Model=model_nodes_HUGO, Mutants = unique(META_mut$Hugo_Symbol), CNA=META_CNA %>% select(-PATIENT_ID) %>% colnames, RNA=META_RNA %>% select(-PATIENT_ID) %>% colnames), filename=NULL, col = "transparent", fill = c("cornflowerblue", "green", "yellow", "darkorchid1"), alpha = 0.5, label.col = c("orange", "white", "darkorchid4", "white", "white", "white", "white", "white", "darkblue", "white", "white", "white", "white", "darkgreen", "white"), cex = 1.5, fontface = "bold", cat.default.pos = "text", cat.col = c("darkblue", "darkgreen", "orange", "darkorchid4"), cat.cex = 1.5, cat.pos = 0, main = "Model genes and available data types"))

```

We investigate relations betwenn RNAseq data and BC subtypes. Subtypes have been defined based on PAM50 method. First, here is the distribution of BC subtypes across the cohort:

```{r PAM50_META, echo=FALSE}
ggplot(META_clin)+geom_bar(aes(x=PAM50, fill=PAM50)) + ggtitle("Distribution of PAM50 subtypes in META cohort")
```

RNAseq is projected on PC1/PC2 space (from Principal Component Analysis), from PAM50 genes only (using only 47 out of 50 genes in PAM50 list present in RNAseq)

```{r PAM50_PCA_META, echo=FALSE, warning=FALSE}

pca_data <- META_RNA %>% select(intersect(PAM50$Gene,colnames(.))) %>% prcomp %>% .$x %>% as.tibble %>%
  mutate(PATIENT_ID=substr(META_RNA$PATIENT_ID,1,12)) %>% full_join(META_clin, by="PATIENT_ID")

ggplot(pca_data, aes(x=PC1, y=PC2, colour=PAM50))+geom_point()+ggtitle("PAM50-Subtype and PAM50-genes PCA projection (based on RNAseq)")

```

# Processing pipeline

## Mutations profiles

We need to assign Boolean effects to mutations: either 0 (inactivating) or 1 (activating). A mutation can stay unassigned in absence of any evidence.

```{r Classical, echo=FALSE, message=FALSE}

Inactivating_mutations <- c("Nonsense_Mutation", "Frame_Shift_Del", "Frame_Shift_Ins", "Splice_Site", "Translation_Start_Site")
#Possible to be less stringent in assignments uncommenting the following line
#Inactivating_mutations <- c("Nonsense_Mutation", "Nonstop_Mutation", "Frame_Shift_Del", "Frame_Shift_Ins", "Splice_Site", "Splice_Region", "Translation_Start_Site")

#OncoKB database for missense assignments
OncoKB <- read_delim("Data/Common/allAnnotatedVariants.txt", delim = "\t") %>% mutate(Label=paste(Gene, Alteration, sep="_"))

#Functionnal inference based on Tokheim et al., 2020+ method
OncoTSG <- read_delim("Data/Common/2020_pancancer.csv", delim = ";", skip = 1) %>% select(gene,`oncogene q-value`,`tsg q-value`)
oncogenes <- OncoTSG %>% filter(`oncogene q-value`<=0.1) %>% select(gene) %>% unlist %>% unname
tsg <- OncoTSG %>% filter(`tsg q-value`<=0.1) %>% select(gene) %>% unlist %>% unname

#The following function performs mutations (from exome-sequencing) assignemnt to 0 or 1 with several methods
mutations_assignment <- function(x){
  output <-  x %>% 
    
    #Assign inactivating mutations to 0 and write label for further missense annotation
    mutate(Label=paste(Hugo_Symbol, `MA:protein.change`, sep="_"), Method=ifelse(Variant_Classification %in% Inactivating_mutations, "Inactivating", NA), BIN=ifelse(Variant_Classification %in% Inactivating_mutations, 0, NA)) %>%
    mutate(Method=ifelse(Variant_Classification == "Silent", "Silent", Method)) %>%
    
    #Assign missense mutations present in OncoKB
    mutate(Method=ifelse(Label %in% OncoKB$Label, "OncoKB", Method)) %>%
    mutate(BIN=ifelse(Method=="OncoKB", 
                      ifelse(grepl("Gain",OncoKB$`Mutation Effect`[match(Label, OncoKB$Label)]),
                             1,
                             ifelse(grepl("Loss",OncoKB$`Mutation Effect`[match(Label, OncoKB$Label)]),
                                    0,
                                    BIN)
                             ),
                      BIN)) %>%
  
  #Use Oncogene/TSG assignments
  mutate(Method=ifelse(is.na(Method) & (Hugo_Symbol %in% oncogenes | Hugo_Symbol %in% tsg), "2020+", Method)) %>%
  mutate(BIN=ifelse(Method=="2020+" & (Hugo_Symbol %in% oncogenes) & (grepl("damaging",PolyPhen) | grepl("deleterious",SIFT)),
                    1,
                    BIN)) %>%
  mutate(BIN=ifelse(Method=="2020+" & (Hugo_Symbol %in% tsg) & (grepl("damaging",PolyPhen) | grepl("deleterious",SIFT)),
                    0,
                    BIN)) %>%
  
  #Write remaining categories
  mutate(Method = ifelse(is.na(Method), "Not Processed", Method)) %>%
  mutate(BIN = ifelse(is.na(BIN), "Not Assigned", BIN)) %>%
  mutate(Status=ifelse(Hugo_Symbol %in% model_nodes_HUGO, "Model-related gene", "Out of scope"))
}

META_mut <- mutations_assignment(META_mut)
METAmodel_mut <- META_mut %>% filter(Hugo_Symbol %in% model_nodes_HUGO)

```

Assignement methods and their respective influence:

```{r Sankey_plots, echo=FALSE, message=FALSE}
sankey_plot <- function(x) {
  
  x <- bind_rows(select(x,Method,BIN) %>% rename(source=Method, target=BIN),select(x,BIN,Status) %>% rename(source=BIN, target=Status)) %>% table %>% as.data.frame %>% as.tibble %>% rename(value=Freq) %>%  filter(value!=0) %>% mutate(source=as.character(source), target=as.character(target)) 
  
  
  # Create nodes DF with the unique sources & targets from input
  nodes <- data.frame(unique(c(x$source,x$target)),stringsAsFactors=FALSE)

  nodes$ID <- as.numeric(rownames(nodes)) - 1 # zero-indexed IDs for sankeyNetwork
  names(nodes) <- c("name", "ID")
  nodes <- nodes %>% as.tibble %>% mutate(NodeGroup=ifelse(name %in% c("0", "1", "Not Assigned"),"Assignment",ifelse(name %in% c("Model-related gene", "Out of scope"),"Model","Method")))

  links <- inner_join(x, nodes, by = c("source"="name")) %>%
    rename(source_ID = ID) %>%
    inner_join(nodes, by = c("target"="name")) %>%
    rename(target_ID = ID) 

  # Create Sankey Plot
  sank <- sankeyNetwork(
    Links = links, Nodes = nodes, Source = "source_ID", Target = "target_ID", Value = "value", NodeID = "name", NodeGroup = "NodeGroup", units = "mutations", fontSize = 12, nodeWidth = 30
  )
  return(sank)
}

print("Sankey plots of mutation assignments depending on methods used")
sankey_plot(META_mut)
print("Sankey plots of mutation assignments depending on methods used (restricted to model-related nodes)")
METAmodel_mut %>% sankey_plot
```

Now we can summarize patient mutation profiles after processing. In following plots we focus on model-related genes only.

```{r Mutations_profiles, echo=FALSE, message=FALSE, warning=FALSE}
#An additional processing steo is required since one gene can carry several mutation
mutation_fusion <- function(x){
  #First, we only retain distinct rows
  output_profile <- x %>% select(PATIENT_ID, Hugo_Symbol, BIN) %>% mutate(BIN=ifelse(BIN=="Not Assigned",NA,BIN)) %>% type_convert %>% distinct
  #Then, when 0&NA or 1&NA cases appear we keep only 0/1 and discard NA information
  logi <- duplicated(output_profile[,1:2]) | duplicated(output_profile[,1:2], fromLast = T)
  output_profile <- filter(output_profile, !(logi & is.na(BIN)))
  #Then, when 0&1 cases appear we keep only 0 since inactivating mutations (usually truncating) are likely to erase gain of function effects (usally achieved through missense)
  logi <- duplicated(output_profile[,1:2]) | duplicated(output_profile[,1:2], fromLast = T)
  output_profile <- filter(output_profile, !(logi & BIN==1))
  
  return(output_profile)
}

META_mut_prof <- META_mut %>% mutation_fusion %>% spread(Hugo_Symbol,BIN) %>% tbl_to_df
#Restrain genes
METAmodel_mut_prof <- META_mut_prof %>% select(one_of(model_nodes_HUGO))

#Plots per genes
METAmodel_mut_prof %>% is.na %>% `!` %>% colSums %>% sort(decreasing=TRUE) %>% head(n=30) %>% barplot(las=2, ylab="# of mutations across cohort", main="Distribution of mutations per genes in META cohort", sub="30 most mutated genes")
#Plots per patients
METAmodel_mut_prof %>% is.na %>% `!` %>% rowSums %>% sort(decreasing=TRUE) %>% plot(ylab="# of mutations", xlab="Patients", main="Distribution of assigned mutations per patients in META cohort")

```

## CNA profiles

For CNA, we have decided to focus on stringent amplifications/deletions corresponding to +2/-2 GISTIC results. We produce the same kind of plots.


```{r CNA_profiles, echo=FALSE, message=FALSE, warning=FALSE}

META_bin <- META_CNA %>% select(PATIENT_ID, one_of(model_nodes_HUGO)) %>% tbl_to_df %>% abs %>% equals(2)

#Plots per genes
META_bin %>% colSums %>% sort(decreasing=TRUE) %>% head(n=30) %>% barplot(las=2, ylab="# of CNA across cohort", main="Distribution of CNA per genes in META cohort", sub="30 most altered genes")

#Plots per patients
META_bin %>% rowSums %>% sort(decreasing=TRUE) %>% plot(ylab="# of CNA", xlab="Patients", main="Distribution of assigned CNA per patients in META cohort")

#Profiles
cna_profile <- function(x){
  interm <- x %>% tbl_to_df
  interm[interm==-1 | interm==0 | interm==1] <- NA
  interm[interm==-2] <- 0
  interm[interm==2] <- 1
  return(interm)
}

META_CNA_prof <- cna_profile(META_CNA)
METAmodel_CNA_prof <- META_CNA_prof %>% select(one_of(model_nodes_HUGO))

```

## RNA profiles

RNA data is intrinsically continuous and therefore require preliminary data processing. It is important to notice that METABRIC data comes from microarray.

### Binarization with classification tree

```{r binarization_normalization_utilities, echo=FALSE, message=FALSE, warning=FALSE}
#function to compute the Bimodality Index (BI) described in Wang et al. (2009)
BI <- function(dataset) {
  x <- dataset
    mc <- Mclust(na.omit(x), G = 2, modelNames = "E", verbose = FALSE)
    if (is.null(mc)) {
      bi <- NA
    } else {
      sigma <- sqrt(mc$parameters$variance$sigmasq)
      delta <- abs(diff(mc$parameters$mean))/sigma
      pi <- mc$parameters$pro[1]
      bi <- delta * sqrt(pi*(1-pi))
    }
  bi
}

#function to binarize the tails of the distribution, based on inter-quartile range (IQR), similar to methods described in teh outlier-sum statistic (Tibshirani and Hastie, 2007). Can be called with a reference dataset
OSclass <- function(exp_dataset, ref_dataset=exp_dataset) {
  classif <-rep(NA,length(exp_dataset))
  q25 <- quantile(ref_dataset,0.25, na.rm = T)
  q75 <- quantile(ref_dataset,0.75, na.rm = T)
  IQR <- q75 - q25 #InterQuartile Range
  classif[exp_dataset>IQR+q75] <- 1
  classif[exp_dataset<q25-IQR] <- 0
  return(classif)
}

#function to to binarize bimodal distributions based on a 2-modes gaussian mixture model (with equal variances). Can be called with a reference dataset
BIMclass <- function(exp_dataset, ref_dataset=exp_dataset) {
  mc <- Mclust(ref_dataset, modelNames = "E", G=2, verbose = FALSE)
  classif <- rep(NA,length(exp_dataset))
  if (diff(mc$parameters$mean)>0){
    thresh_down <- max(mc$data[mc$classification==1 & mc$uncertainty <= 0.05])
    thresh_up <- min(mc$data[mc$classification==2 & mc$uncertainty <= 0.05])
    classif[exp_dataset<=thresh_down] <- 0
    classif[exp_dataset>=thresh_up] <- 1
  } else if (diff(mc$parameters$mean)<0){
    thresh_down <- max(mc$data[mc$classification==2 & mc$uncertainty <= 0.05])
    thresh_up <- min(mc$data[mc$classification==1 & mc$uncertainty <= 0.05])
    classif[exp_dataset<=thresh_down] <- 0
    classif[exp_dataset>=thresh_up] <- 1
  }
  return(classif)
}

#function for normalization of zero-inflated data
norm_fun_lin <- function(xdat, reference = xdat){
  x_proc <- (xdat-quantile(reference, 0.01))/quantile(xdat-quantile(reference, 0.01), 0.99)
  x_proc[x_proc<0] <- 0
  x_proc[x_proc>1] <- 1
  x_proc
}

#function for normalization of unimodal data
norm_fun_sig <- function(xdat, reference = xdat){
  xdat <- xdat - median(reference, na.rm = T)
  lambda <- log(3)/mad(reference, na.rm = T)
  transformation <- function(x){
    y <- 1/(1+exp(-lambda*x))
    y
  }
  transformation(xdat) 
}

#function for normalization of unimodal data
norm_fun_bim <- function(xdat, reference = xdat) {
  mc <- Mclust(reference, modelNames = "E", G=2, verbose = FALSE)
  pred <- predict.Mclust(mc,xdat)
  if (diff(mc$parameters$mean)>0){
    normalization <- pred$z[,2]
  } else if (diff(mc$parameters$mean)<0){
    normalization <- pred$z[,1]
  }
  normalization
}
```

```{r RNA_profiles_original_criteria, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE}

#Here we compute all statistical tools and criteria needed to perform the classification of distributions in the following categories: discarded, zero-inflated, unimodal and bimodal
compute_criteria <- function(exp_dataset){
  exp_dataset <- exp_dataset %>% select(-PATIENT_ID)
  criteria <- tibble(Gene=colnames(exp_dataset), Dip=NA, BI=NA, Kurtosis=NA, DropOutRate=NA, MeanNZ=NA, DenPeak=NA, Amplitude=NA)
  
  #Compute
  pb = txtProgressBar(min = 1, max = ncol(exp_dataset), initial = 1) 
  for (i in 1:ncol(exp_dataset)){
    x <- na.omit(unlist(exp_dataset[,i]))
    criteria$Amplitude[i] <- max(x)-min(x)
    
    if (criteria$Amplitude[i] !=0){
      criteria$Dip[i] <- dip.test(x)$p.value
      criteria$BI[i] <- BI(x)
      criteria$Kurtosis[i] <- kurtosis(x)-3
      criteria$DropOutRate[i] <- sum(x==0)/length(x)
      criteria$MeanNZ[i] <- sum(x)/sum(x!=0)
      den <- density(x, na.rm = T)
      criteria$DenPeak[i] <- den$x[which.max(den$y)]
    }
    
    setTxtProgressBar(pb,i)
  }
  
  threshold <- median(criteria$Amplitude)/10
  criteria <- criteria %>% 
    mutate(Category=ifelse(Amplitude<threshold | DropOutRate>0.95, "Discarded", NA)) %>%
    mutate(Category=ifelse(is.na(Category) & (BI>1.5 & Dip<0.05 & Kurtosis < 1),"Bimodal",Category)) %>%
    mutate(Category=ifelse(is.na(Category) & DenPeak<threshold, "ZeroInf", Category)) %>%
    mutate(Category=ifelse(is.na(Category), "Unimodal", Category))
  
  return(criteria)
}

criteria_META <- compute_criteria(META_RNA)

```

Now, what about the distribution of gene categories (Bimodal, Unimodal...) across the cohort?

```{r RNA_profiles_original_catDistrib, echo=FALSE, message=FALSE, warning=FALSE}
print("META assignments:")
kable(t(table(criteria_META$Category)))

print("META assignments for model-related nodes:")
kable(t(table(criteria_META %>% filter(Gene %in% model_nodes_HUGO) %>% select(Category))))

```

Here are some distributions plots randomly picked in each category in META cohort

```{r RNA_profiles_original_catPlot, echo=FALSE, message=FALSE, warning=FALSE}
criteria_META %>% filter(Category=="Bimodal") %>% select(Gene) %>% slice(sample(nrow(.),5)) %>% unlist %>% select(META_RNA,.) %>%  gather %>%
  ggplot(mapping = aes(x = value)) + geom_histogram(bins = 30) + facet_wrap(~key, scales = 'free') + ggtitle("5 random Bimodal genes")

criteria_META %>% filter(Category=="Unimodal") %>% select(Gene) %>% slice(sample(nrow(.),20)) %>% unlist %>% select(META_RNA,.) %>%  gather %>%
  ggplot(mapping = aes(x = value)) + geom_histogram(bins = 30) + facet_wrap(~key, scales = 'free') + ggtitle("20 random Unimodal genes")

```

And depending on distribution category, we can perform binarization

```{r RNA_profiles_original_profiles, echo=FALSE, message=FALSE, warning=FALSE}
#function to apply the proper binarization method depending on the gene expression distribution category
binarize_exp <-  function(exp_dataset, ref_dataset, ref_criteria, gene, show.plot=F){
  if(!missing(gene)){
    
    gene_cat <- ref_criteria %>% filter(Gene==gene) %>% select(Category) %>% unlist
    x <- unlist(select(exp_dataset,gene))
    x_ref <- unlist(select(ref_dataset,gene))
    
    if (gene_cat=="Discarded"){
        stop("Discarded gene")
      
      } else if (gene_cat=="Bimodal"){
        gene_bin <- BIMclass(x,x_ref)
        
      } else {
        gene_bin <- OSclass(x,x_ref)
      }
    names(gene_bin) <- exp_dataset$PATIENT_ID
    if(show.plot==T){
      if(all(is.na(gene_bin))){
        tibble(Continuous=x) %>% ggplot(aes(x=Continuous))+geom_histogram(bins=30)+ggtitle(gene)
      } else {
        tibble(Continuous=x, Discrete=factor(gene_bin)) %>% ggplot(aes(x=Continuous, fill=Discrete))+geom_histogram(bins=30)+ggtitle(gene)
      }
    } else {
      return(gene_bin)
    }
    
  } else {
    exp_dataset <- tbl_to_df(exp_dataset) 
    ref_dataset <- tbl_to_df(ref_dataset)
    if(dim(exp_dataset)[2] != dim(ref_criteria)[1]){stop("Different number of genes")}
    logi_dis <- ref_criteria$Category=="Discarded"
    logi_OS <- ref_criteria$Category=="Unimodal" | ref_criteria$Category=="ZeroInf"
    logi_bim <- ref_criteria$Category=="Bimodal"
    exp_dataset[,logi_dis] <- lapply(exp_dataset[,logi_dis], function(x) rep(NA, length(x)))
    exp_dataset[,logi_OS] <- mapply(function(x,y) OSclass(x,y), exp_dataset[,logi_OS], ref_dataset[,logi_OS])
    exp_dataset[,logi_bim] <- mapply(function(x,y) BIMclass(x,y), exp_dataset[,logi_bim], ref_dataset[,logi_bim])
    
    return(exp_dataset)
  }
  
}

print("Bimodal example:")
binarize_exp(META_RNA,META_RNA, criteria_META, "ESR1", T)
print("Unimodal example:")
binarize_exp(META_RNA,META_RNA, criteria_META, "FOXO3", T)



META_RNA_prof <- binarize_exp(META_RNA,META_RNA, criteria_META)

META_RNA_prof %>% select(one_of(model_nodes_HUGO)) %>% is.na %>% `!` %>% rowSums %>% sort(decreasing=TRUE) %>% plot(ylab="# of bin. RNA", xlab="Patients", main="Distribution of binarized RNA per patients in META cohort")

META_RNA_prof %>% select(one_of(model_nodes_HUGO)) %>% is.na %>% `!` %>% colSums %>% sort(decreasing=TRUE) %>%  head(n=50) %>% barplot(las=2, ylab="# of mutations across cohort", main="Distribution of mutations for top 50 genes in META cohort")

#Profiles
METAmodel_RNA_prof <- META_RNA_prof %>% select(one_of(model_nodes_HUGO))
```

### Normalization

```{r norm_profiles, echo=FALSE, message=FALSE, warning=FALSE}

normalize_exp <-  function(exp_dataset, ref_dataset, ref_criteria, gene, show.plot=F){
  if(!missing(gene)){
    
    gene_cat <- ref_criteria %>% filter(Gene==gene) %>% select(Category) %>% unlist
    x <- unlist(select(exp_dataset,gene))
    x_ref <- unlist(select(ref_dataset,gene))
    
    if (gene_cat=="Discarded"){
        stop("Discarded gene")
      
      } else if (gene_cat=="Bimodal"){
        gene_bin <- norm_fun_bim(x,x_ref)
        
      } else if (gene_cat=="Unimodal"){
        gene_bin <- norm_fun_sig(x,x_ref)
        
      } else {
        gene_bin <- norm_fun_lin(x,x_ref)
      }
    names(gene_bin) <- exp_dataset$PATIENT_ID
    
    if(show.plot==T){
      gene_bin %>% unlist %>% as.data.frame %>% ggplot(aes(x=.)) + geom_histogram(bins=30)+xlab(gene)
    } else {
      return(gene_bin)
    }
    
  } else {
    exp_dataset <- tbl_to_df(exp_dataset) 
    ref_dataset <- tbl_to_df(ref_dataset)
    if(dim(exp_dataset)[2] != dim(ref_criteria)[1]){stop("Different number of genes")}
    logi_dis <- ref_criteria$Category=="Discarded"
    logi_uni <- ref_criteria$Category=="Unimodal"
    logi_zero <- ref_criteria$Category=="ZeroInf"
    logi_bim <- ref_criteria$Category=="Bimodal"
    exp_dataset[,logi_dis] <- lapply(exp_dataset[,logi_dis], function(x) rep(NA, length(x)))
    exp_dataset[,logi_uni] <- mapply(function(x,y) norm_fun_sig(x,y), exp_dataset[,logi_uni], ref_dataset[,logi_uni])
    exp_dataset[,logi_zero] <- mapply(function(x,y) norm_fun_lin(x,y), exp_dataset[,logi_zero], ref_dataset[,logi_zero])
    exp_dataset[,logi_bim] <- mapply(function(x,y) norm_fun_bim(x,y), exp_dataset[,logi_bim], ref_dataset[,logi_bim])
    
    return(exp_dataset)
  }
  
}

META_RNA_prof_norm <- normalize_exp(META_RNA, META_RNA, criteria_META)

METAmodel_RNA_prof_norm <- META_RNA_prof_norm %>% select(one_of(model_nodes_HUGO))
```

# Merged profiles

```{r merge_patients_genes, echo=FALSE, message=FALSE, warning=FALSE}

common_patients_MCR <- Reduce(intersect, list(rownames(METAmodel_mut_prof),rownames(METAmodel_CNA_prof),rownames(METAmodel_RNA_prof)))

common_genes_MCR <- Reduce(intersect, list(colnames(METAmodel_CNA_prof),colnames(METAmodel_RNA_prof)))

```

## Data types relations

Before merging independant profiles into multi-omics profiles, let's have a look at relations between data types

```{r Relations, echo=FALSE, message=FALSE, warning=FALSE}
common_genes <- Reduce(intersect,list(colnames(META_mut_prof),colnames(META_CNA_prof),colnames(META_RNA_prof)))
common_patients <- Reduce(intersect,list(rownames(META_mut_prof),rownames(META_CNA_prof),rownames(META_RNA_prof)))

plot_data <- list(setNames(META_mut_prof[common_patients,],paste0('mut.',names(META_mut_prof))) %>% sapply(as.factor), 
                  setNames(META_CNA_prof[common_patients,],paste0('CNA.',names(META_CNA_prof))) %>% sapply(as.factor), 
                  setNames(META_RNA %>% filter(PATIENT_ID %in% common_patients) %>% select(common_genes),paste0('RNA.',common_genes))) %>%
  cbind.data.frame

plot_mapping <- function(plot_data, gene){
  
  if(paste0("RNA.",gene) %in% colnames(plot_data) & paste0("mut.",gene) %in% colnames(plot_data)){
    plot_up <- plot_data %>% ggplot(aes_string(x=paste0("RNA.", gene), fill=paste0("mut.", gene)))+geom_histogram(bins=30)
  } else {
    print("No RNA/mut plot")
  }
  
  if(paste0("RNA.",gene) %in% colnames(plot_data) & paste0("CNA.",gene) %in% colnames(plot_data)){
    plot_down <- plot_data %>% ggplot(aes_string(x=paste0("RNA.", gene), fill=paste0("CNA.", gene)))+geom_histogram(bins=30)
  } else {
    print("Not RNA/CNA plot")
  }
  
  grid.arrange(plot_up, plot_down)
}

# Define a set of genes for which there are values on each of the plotted datasets. If the gene only has NAs on one dataset, the plot will show an error.
plot_mapping(plot_data,"AKT1")
plot_mapping(plot_data,"TP53")

#Compute CNA/RNA correlation in META
common_genes <- intersect(colnames(select(META_CNA,-PATIENT_ID)),colnames(select(META_RNA,-PATIENT_ID)))
common_patients <- intersect(META_CNA$PATIENT_ID,META_RNA$PATIENT_ID)
ken_META_CNA_RNA <- mapply(function(x,y) cor(x,y,method = "kendall"), tbl_to_df(META_CNA)[common_patients,common_genes], tbl_to_df(META_RNA)[common_patients,common_genes])
ken_META_CNA_RNA %>% as.data.frame %>% ggplot(aes(x=.)) + geom_histogram(bins=30) + geom_vline(xintercept = median(ken_META_CNA_RNA, na.rm = T)) + ggtitle("CNA/RNA correlation distribution for each gene across cohort (with median)")


```

### Mutations and CNA

In particular, is there any mutation/CNA binary inconsistency?

```{r incons_mut_CNA, echo=FALSE, message=FALSE, warning=FALSE}
common_genes <- intersect(colnames(METAmodel_CNA_prof), colnames(METAmodel_mut_prof))
common_patients <- intersect(rownames(METAmodel_CNA_prof), rownames(METAmodel_mut_prof))

incons <- which(METAmodel_mut_prof[common_patients, common_genes]==0 & METAmodel_CNA_prof[common_patients, common_genes]==1 | METAmodel_mut_prof[common_patients, common_genes]==1 & METAmodel_CNA_prof[common_patients, common_genes]==0, arr.ind = T) %>% as.data.frame %>% mutate(Patient=common_patients[row],Gene=common_genes[col], Mut=diag(as.matrix(METAmodel_mut_prof[Patient,Gene])), CNA=diag(as.matrix(METAmodel_CNA_prof[Patient,Gene]))) %>% select(Patient,Gene, Mut, CNA)

if (nrow(incons)==0){
  print("No mutation/CNA binary inconsistencies")
}else{
  kable(incons)
}
```

In case of ambiguity, pritority is given to mutations over CNA

```{r merge_mut_CNA, echo=FALSE, message=FALSE, warning=FALSE}

mut_plus_CNA <- function(mut,cna){
  merged <- mut
  merged[is.na(merged)] <- cna[is.na(merged)]
  return(merged)
}

METAmodel_mutCNA_prof = METAmodel_mutCNAneg_prof <- matrix(nrow = length(common_patients_MCR), ncol = dim(METAmodel_CNA_prof)[2], dimnames = list(common_patients_MCR,colnames(METAmodel_CNA_prof))) %>% as.data.frame
METAmodel_CNAneg_prof <- METAmodel_CNA_prof
METAmodel_CNAneg_prof[METAmodel_CNAneg_prof==1] <- NA

for(i in 1:dim(METAmodel_mutCNA_prof)[2]){
  gene <- colnames(METAmodel_mutCNA_prof)[i]
  if (gene %in% colnames(METAmodel_mut_prof)){
    METAmodel_mutCNA_prof[,gene] <- mut_plus_CNA(METAmodel_mut_prof[common_patients_MCR,gene], METAmodel_CNA_prof[common_patients_MCR,gene])
    METAmodel_mutCNAneg_prof[,gene] <- mut_plus_CNA(METAmodel_mut_prof[common_patients_MCR,gene], METAmodel_CNAneg_prof[common_patients_MCR,gene])
  } else {
    METAmodel_mutCNA_prof[,gene] <- METAmodel_CNA_prof[common_patients_MCR,gene]
    METAmodel_mutCNAneg_prof[,gene] <- METAmodel_CNAneg_prof[common_patients_MCR,gene]
  }
}


```

### CNA and RNA

```{r incons_CNA_RNA, echo=FALSE, message=FALSE, warning=FALSE}
common_genes <- intersect(colnames(METAmodel_CNA_prof), colnames(METAmodel_RNA_prof))
common_patients <- intersect(rownames(METAmodel_CNA_prof), rownames(METAmodel_RNA_prof))

incons <- which(METAmodel_CNA_prof[common_patients, common_genes]==0 & METAmodel_RNA_prof[common_patients, common_genes]==1 | METAmodel_CNA_prof[common_patients, common_genes]==1 & METAmodel_RNA_prof[common_patients, common_genes]==0, arr.ind = T) %>% as.data.frame %>% mutate(Patient=common_patients[row],Gene=common_genes[col],  CNA=diag(as.matrix(METAmodel_CNA_prof[Patient,Gene])), RNA=diag(as.matrix(METAmodel_RNA_prof[Patient,Gene]))) %>% select(Patient,Gene,CNA, RNA)

if (nrow(incons)==0){
  print("No RNA/CNA binary inconsistencies")
}else{
  kable(incons)
}

```

In case of ambiguity, pritority is given to RNA over CNA

```{r merge_CNA_RNA, echo=FALSE, message=FALSE, warning=FALSE}

CNA_plus_RNA <- function(cna,rna){
  merged <- rna
  merged[is.na(merged)] <- cna[is.na(merged)]
  return(merged)
}

METAmodel_mutCNA_prof <- matrix(nrow = length(common_patients_MCR), ncol = dim(METAmodel_CNA_prof)[2], dimnames = list(common_patients_MCR,colnames(METAmodel_CNA_prof))) %>% as.data.frame

for(i in 1:dim(METAmodel_mutCNA_prof)[2]){
  gene <- colnames(METAmodel_mutCNA_prof)[i]
  if (gene %in% colnames(METAmodel_mut_prof)){
    METAmodel_mutCNA_prof[,gene] <- mut_plus_CNA(METAmodel_mut_prof[common_patients_MCR,gene], METAmodel_CNA_prof[common_patients_MCR,gene])
  } else {
    METAmodel_mutCNA_prof[,gene] <- METAmodel_CNA_prof[common_patients_MCR,gene]
  }
}


```


### Mut and RNA

```{r mut_RNA, echo=FALSE, message=FALSE, warning=FALSE}
common_genes <- intersect(colnames(METAmodel_mut_prof), colnames(METAmodel_RNA_prof))
common_patients <- intersect(rownames(METAmodel_mut_prof), rownames(METAmodel_RNA_prof))

incons <- which(METAmodel_mut_prof[common_patients, common_genes]==0 & METAmodel_RNA_prof[common_patients, common_genes]==1 | METAmodel_mut_prof[common_patients, common_genes]==1 & METAmodel_RNA_prof[common_patients, common_genes]==0, arr.ind = T) %>% as.data.frame %>% mutate(Patient=common_patients[row],Gene=common_genes[col],  Mut=diag(as.matrix(METAmodel_mut_prof[Patient,Gene])), RNA=diag(as.matrix(METAmodel_RNA_prof[Patient,Gene]))) %>% select(Patient,Gene,Mut,RNA)

if (nrow(incons)==0){
  print("No mutation/RNA binary inconsistencies")
}else{
  kable(incons)
}

# Define a set of genes for which there are values on each of the plotted datasets.
plot_mapping(plot_data,'PIK3CA')

mut_plus_RNA <- function(mutations,expression){
  all_genes <- union(colnames(mutations), colnames(expression))
  common_patients <- intersect(rownames(mutations), rownames(expression))
  merged_matrix <- matrix(nrow = length(common_patients), ncol = length(all_genes), dimnames = list(common_patients,all_genes)) %>% as.data.frame
  
  mut_plus_exp_vec <- function(mut,exp){
    merged <- mut
    #merged[merged==1 & exp==0] <- 0
    merged[is.na(merged)] <- exp[is.na(merged)]
    return(merged)
  }
  
  for(gene in colnames(merged_matrix)){
    if ((gene %in% colnames(mutations)) & (gene %in% colnames(expression))){
      merged_matrix[,gene] <- mut_plus_exp_vec(mutations[common_patients,gene], expression[common_patients,gene])
    } else if (gene %in% colnames(mutations)) {
      merged_matrix[,gene] <- mutations[common_patients,gene]
    } else {
      merged_matrix[,gene] <- expression[common_patients,gene]
    }
  }
  return(merged_matrix)
}

METAmodel_mutRNA_prof <- mut_plus_RNA(METAmodel_mut_prof, METAmodel_RNA_prof)

```


## Write profiles

```{r write, echo=FALSE, message=FALSE, warning=FALSE}

inverse_mapping <- function(dataset, datatype = 'bin', geneindex_var = data.frame(lapply(geneindex, as.character), stringsAsFactors=FALSE)){
  patients <- rownames(dataset)
  
  dupli <- geneindex_var %>% filter(duplicated(geneindex_var$V2))
  if (nrow(dupli)>0){
    for (i in 1:nrow(dupli)){
      gene_dupli <- dupli[i,2]
      node_dupli <- dupli[i,1]
      new_gene <- paste(gene_dupli,node_dupli,sep="_")
      index_dupli <- match_df(geneindex_var,dupli[1,]) %>% rownames %>% as.numeric
      geneindex_var[index_dupli,2] <- new_gene
      if (gene_dupli %in% colnames(dataset)){
        dataset[,new_gene] <- dataset[,gene_dupli]
      }
    }
  }
  
  geneindex_corres <- geneindex_var %>% column_to_rownames(var="V2")
  
  erased_genes <- list()
  for (gene in colnames(dataset)){
    #print(gene)
    if (!(gene %in% erased_genes)){
      index <- match(gene, colnames(dataset))
      node <- geneindex_corres[gene,] %>% as.character
      family_size <- geneindex_var %>% filter(V1==node) %>% nrow
      if (family_size==0){
        stop("Correspondency problem")
      } else if (family_size==1){
        colnames(dataset)[index] <- node
      } else if (family_size>1){
        family_members <- geneindex_var %>% filter(V1==node) %>% select(V2) %>% unlist %>% as.character
        family_df <- dataset %>% select(one_of(family_members)) #%>% select_if(~sum(!is.na(.)) > 0)
        if (ncol(family_df)==1){
          new_column <- family_df %>% unlist
          erased_name <- colnames(family_df)
          dataset <- select(dataset, -one_of(erased_name))
        } else if (ncol(family_df)>1 & datatype=='bin'){
          interm <- rowMeans(family_df, na.rm = T)
          interm[interm!=0 & interm !=1] <- NA
          interm[is.nan(interm)] <- NA
          new_column <- interm
          remove_index <- match(family_members, colnames(dataset))
          dataset <- select(dataset, -remove_index)
          erased_genes <- c(erased_genes, family_members) %>% unlist
        } else if (ncol(family_df)>1 & datatype=='norm'){
          new_column <- rowMeans(family_df, na.rm = T)
          remove_index <- match(family_members, colnames(dataset))
          dataset <- select(dataset, -remove_index)
          erased_genes <- c(erased_genes, family_members) %>% unlist
        }
        dataset <- mutate(dataset, merge=new_column)
        colnames(dataset)[ncol(dataset)] <- node
        
      }
    }
    
  }
  rownames(dataset) <- patients
  return(dataset)
}

if (!dir.exists("Results/Profiles")){
  dir.create("Results/Profiles")
}

#METAmodel_mut_prof_red <- METAmodel_mut_prof[,colSums(!is.na(METAmodel_mut_prof)) > 0]
write.csv(inverse_mapping(METAmodel_mut_prof),"Results/Profiles/Zanudo_META_mutations.csv")

write.csv(inverse_mapping(METAmodel_CNA_prof),"Results/Profiles/Zanudo_META_CNA.csv")

write.csv(inverse_mapping(METAmodel_mutCNA_prof),"Results/Profiles/Zanudo_META_mutCNA.csv")
write.csv(inverse_mapping(METAmodel_mutCNAneg_prof),"Results/Profiles/Zanudo_META_mutCNAneg.csv")

write.csv(inverse_mapping(METAmodel_RNA_prof),"Results/Profiles/Zanudo_META_RNA.csv")
write.csv(inverse_mapping(METAmodel_RNA_prof_norm, datatype='norm'),"Results/Profiles/Zanudo_META_RNA_norm.csv")

write.csv(inverse_mapping(METAmodel_mutRNA_prof),"Results/Profiles/Zanudo_META_mutRNA.csv")


```

